---
title: "ggplot for Time Series Data"
author: "Matthew Heino"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE)
```


# Data Description

The data is from the CSV file with the file name "Crime3years.rds". The data represents the crime rate by state over a period of years.  The range of years is from 2007 to 2018. The data set is composed of three columns or variables. The columns are the following:

* Variable 1: **Data** - (dttm) holds the date in the following format: 2017-09-01 00:00:00
* Variable 2: **CrimeRate** - (int) the crime rate

# Data Wrangling 

The data needed to be prepared before it could be processed. The steps are the following:

* Step 1: Read in the data from the rds file ("Crime3years.rds").
* Step 2: View the contents of the data frame.
* Step 3: Select two years from the data set
* Step 4: Create the different parts of the date
* Step 5:   
* Step 6: 
* Step 7: 

Loading the appropriate packages will be necessary.

### Load and Install the Required Packages

The required packages for this assignment were:

* tidyverse
* zoo

```{r}

if(!require(tidyverse)) {install.packages('tidyverse')}; require(tidyverse)
if(!require(zoo)) {install.packages('zoo')}; require(zoo)

library(tidyverse)
library(zoo)

```

#### Set the working directory.

```{r}

setwd(dirname(rstudioapi::getSourceEditorContext()$path))

```

###  Read in the data from the rds file ("Crime3years.rds")

```{r}

crime_bat <- readRDS("Crime3years.rds")

```

### Step 2: View the contents of the data frame

```{r}

glimpse(crime_bat)

```

###Step 3: Select two years from the data set
```{r}

two_years = 365*24*2
start_date = 553
end_date = start_date + two_years - 1

new_crime_bat <- crime_bat[start_date:end_date, ]

```

### Step 4: Create the different parts of the date



